An Improved Method for Coupling Hydrodynamics with Reaction Networks and Nuclear Statistical Equilibrium

### 2021-07-13 simplified_sdc_intro/paper.tex

We explored this and other approaches (including not evolving an
energy / temperature equation during the reaction step) in
\citep{strang_rnaas} and showed that the above formulation gets second
order convergence.  However, for very strong reactions, when using
Strang splitting the state can drift significantly off of the smooth
solution to the coupled reactive hydrodynamics equations, as shown
graphically in \cite{astronum:2018} using an earlier version of the
present algorithm.

@@

@@

There is one additional part of the time integration algorithm.
Because $T$ evolves during the reactions (for both Strang and SDC), it
is possible for a zone to start out not in NSE but evolve into NSE
during the reaction update.  When this happens, the ODE integrator may
fail, because an excessive number of timesteps is required.  Rather
than trigger our retry mechanism and throw away the entire timestep
and start over with a smaller $\Delta t$, we first check if the zone
evolved into NSE, and if so, we redo the zone update using the NSE
prescription.

@@

While VODE can compute the entire Jacobian, ${\bf J}$ numerically via differencing, we found
that does not give reliable results.  Instead, we compute $\partial {\bf w}/\partial {\bf U}^\prime$ analytically and $\partial {\bf R}/\partial {\bf w}$ numerically via one-sided differencing.
The resulting Jacobian, as computed via equation~\ref{eq:factored_jac}, is larger than
expected by the integrator because of the inclusion of the $\rho$
terms in the intermediate steps, so we remove the row corresponding to
$R(\rho)$ and the column corresponding to $\partial/\partial \rho$
before passing the Jacobian to VODE.

>>>>

We explored this and other approaches (including not evolving an
energy / temperature equation during the reaction step) in
\citet{strang_rnaas} and showed that the above formulation gets second
order convergence.  However, for very strong reactions, when using
Strang splitting the state can drift significantly off of the smooth
solution to the coupled reactive hydrodynamics equations, as shown
graphically in \cite{astronum:2018} using an earlier version of the
present algorithm.

@@

      {\color{red} We enforce that the species interface states remain in $[0, 1]$ after adding the reactive source.}

@@

There is one additional part of the time integration algorithm.
Because $T$ evolves during the reactions (for both Strang and SDC), it
is possible for a zone to start out not in NSE but evolve into NSE
during the reaction update.  When this happens, the ODE integrator may
fail, because an excessive number of timesteps is required.  Rather
than trigger our retry mechanism and throw away the entire timestep
and start over with a smaller $\Delta t$, we first check if the zone
evolved into NSE, and if so, we redo the zone update using the NSE
prescription.  {\color{red} We optionally apply a relaxation factor, typically, $0.9$, to the NSE conditions to allow a state that is close to NSE after an integrator failure enter NSE.}

@@

While VODE can compute the entire Jacobian, ${\bf J}$ numerically via differencing, we found
that does not give reliable results.  Instead, we compute $\partial {\bf w}/\partial {\bf U}^\prime$ analytically and $\partial {\bf R}/\partial {\bf w}$ numerically via one-sided differencing, following
the algorithm in \cite{lsode} to minimize numerical noise.
The resulting Jacobian, as computed via equation~\ref{eq:factored_jac}, is larger than
expected by the integrator because of the inclusion of the $\rho$
terms in the intermediate steps, so we remove the row corresponding to
$R(\rho)$ and the column corresponding to $\partial/\partial \rho$
before passing the Jacobian to VODE.
